% R bootcamp, Module 10: Advanced topics
% August 2013, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(foreign)
earnings <- read.dta('../data/heights.dta')
```

# Object-oriented programming in R

Note to BB: start recording.

# Working with S3 classes and methods

# Using S4 classes and methods

# A brief mention of Reference Classes


# Computing on the language

One of the powerful capabilities you have in R is the ability to use R to modify and create R code. 

First we need to understand a bit about how R code is stored and manipulated when we don't want to immediately evaluate it.

# Expressions

system.time(a = 5)
system.time({a = 5})

code <- quote(a <- 5)

# Using R to write and evaluate code



# Error and warning messages

When you write your own functions, and particularly for distributing to others, it's a good idea to:

* Check for possible errors (particularly in the input arguments) and give the user an informative error message
* Warn them if you're doing something they might not have anticipated

We can use `stop()` and `warning()` to do this. They're the same functions that are being called when you see an error message or a warning in reaction to your own work in R.

```{r}
mysqrt <- function(x) {
  if(is.list(x)) {
    warning("x is a list; converting to a vector")
    x <- unlist(x)
  }
  if(!is.numeric(x)) {
    stop("What is the square root of 'bob'?")
  } else {
      if(any(x < 0)) {
        warning("mysqrt: found negative values; proceeding anyway")
        x[x < 0] <- NaN
        x[x >= 0] <- (x[x >= 0])^(1/2)
        return(x)
      } else return(x^(1/2))
  }
}

mysqrt(c(1, 2, 3))
mysqrt(c(5, -7))
mysqrt(c('asdf', 'sdf'))
mysqrt(list(5, 3, 'ab'))
sqrt(c(5, -7))
sqrt('asdf')
sqrt(list(5, 3, 2))
```

So we've done something similar to what `sqrt()` actually does in R.

# 'Catching' errors

When you automate analyses, sometimes an R call will fail. But you don't want all of your analyses to grind to a halt because one failed. Rather, you want to catch the error, record that it failed, and move on.

For me this is most critical when I'm doing stratified analyses or sequential operations.

The `try()` function is a powerful tool here.

# Why we need to `try()`

Suppose we tried to do a stratified analysis of earnings on height within education levels. I'm going to do this as a for loop for pedagogical reasons, but again, it would be better to do this with apply/by/plyr type tools.


```{r}
mod <- list()
for(edLevel in unique(earnings$ed)) {
print(edLevel)
            sub <- subset(earnings, ed == edLevel)
            mod[[edLevel]] <- lm(earn ~ height, data = sub)
}
```

1) What happened?
2) Why did it go through the education levels in that order? (12, 16, 17, ...)

# How we can `try()` harder

```{r}
mod <- list()
for(edLevel in unique(earnings$ed)) {
print(edLevel)
            sub <- subset(earnings, ed == edLevel)
            tmp <- try(lm(earn ~ height, data = sub))
            if(is(tmp, "try-error")) mod[[edLevel]] <- NA else mod[[edLevel]] <- tmp
}
mod[[2]]
mod[[3]]
```


# File encodings

Text (either in the form of a file with regular language in it or a data file with fields of character strings) will often contain characters that are not part of the [limited ASCII set of characters](http://en.wikipedia.org/wiki/ASCII), which has 128 characters and control codes; basically what you see on a standard US keyboard.

UTF-8 is an encoding for the Unicode characters that include more than 110,000 characters from 100 different alphabets/scripts. It's widely used on the web.

Latin-1 encodes a small subset of Unicode and contains the characters used in many Europeon languages (e.g., letters with accents).

# Dealing with encodings in R

To read files with other characters correctly into R, you may need to tell R what encoding the file is in. E.g., see help on `read.table()` for the *fileEncoding* and *encoding* arguments. 

With strings already in R, you can convert between encodings:
```{r}
text <- "_Melhore sua seguran\xe7a_"
iconv(text, from = "latin1", to = "UTF-8")
iconv(text, from = "latin1", to = "ASCII", sub = "???")
```

You can mark a string with an encoding so R can display it correctly:
```{r}
x <- "fa\xE7ile"
Encoding(x) <- "latin1"
x
```

# Line endings in text files

Windows, Mac, and Linux handle line endings in text files somewhat differently. So if you read a text file into R that was created in a different operating system you can run into difficulties.

* In Windows lines end in both a newline (the ASCII character \n) and a carriage return (\r). 
* In UNIX, lines end in only a newline.
* On a Mac, lines may end only in a carriage return.

So in UNIX you might see ^M at the end of lines when you open a Windows file in a text editor. The dos2unix or fromdos commands can do the necessary conversion

In Windows you might have a UNIX text file appear to be all one line. The unix2dos or todos commands in UNIX can do the conversion. 

There is a UNIX utility call mac2unix that can convert Mac text files.

There may also be Windows tools to deal with this. 

# Working with databases

R has the capability to read and write from a variety of relational database management systems (DBMS). Basically a database is a collection of rectangular format datasets (tables). Some of these tables have fields in common so it makes sense to merge (i.e., join) information from multiple tables. E.g., you might have a database with a table of student information, a table of teacher information and a table of school information. 

The *DBI* package provides a front-end for manipulating databases from a variety of DBMS (MySQL, SQLite, Oracle, among others)

Basically, you tell the package what DBMS is being used on the back-end, link to the actual database, and then you can use the standard functions in the package regardless of the back-end. 

# Database example

The Current Index to Statistics contains records of article and author information from most Statistics journals. 

The database is not freely available, so it's not on the course bSpace or Github, but I can do a demo here.

```{r}
library(RSQLite)  # DBI is a dependency
db <- dbConnect(SQLite(), dbname = "../data/cis.db") # cis.db is an SQLite database
dbListTables(db)
dbListFields(db, "articles")
breiman <- dbGetQuery(db, "select * from authors where name like 'Breiman%'")
```

# Breakout

Have a snack!
