% R bootcamp, Module 2: Working with objects and data
% August 2013, UC Berkeley
% Chris Paciorek and (add others)

# Matrices and Arrays

Note: remember to start recording.

Let's review matrices

- Matrices are two-dimensional collections of values of the same type
- We can have numeric matrices, character matrices, or logical matrices
- You can't mix types within a matrix
```{r}
mat <- matrix(rnorm(12), nrow = 3, ncol = 4)
arr <- array(rnorm(12), c(2, 3, 4))
mat
arr

mat*4
mat <- cbind(mat, 1:3)

```

# Attributes

Objects have *attributes*.

```{r attr}
attributes(mat)
rownames(mat) <- c('first', 'middle', 'last')
mat
attributes(mat)

mat <- matrix(rnorm(12), nrow = 3, ncol = 4)
mat[4]
attributes(mat)$dim <- NULL
mat
is.matrix(mat)
```
What can you infer about what a matrix is in R?

# Matrices are stored column-major

This is like Fortran but not like C. 

```{r}
mat <- matrix(1:12, 3, 4)
c(mat)
identical(mat, matrix(c(mat), 3, 4))
```
So you can go back and forth smoothly. 


# Missing values and other special values

Since it was designed by statisticians, R handles missing value very well relative to other languages.

* `NA` is a missing value
```{r}
vec <- rnorm(12)
vec[c(3, 5)] <- NA
vec
length(vec)
sum(vec)
sum(vec, na.rm = TRUE)
hist(vec)
is.na(vec)
```
Be careful because many R functions won't warn you that they are ignoring the missing values.

* To infinity and beyond
big <- 1e500 
big
big + 7


* `NaN` stands for Not a Number
```{r}
sqrt(-5)
big - big
1/0
```

* `NULL`
```{r nulldata}
vec <- c(vec, NULL) 
vec
length(vec)
a <- NULL
a + 7
a[3, 4]
is.null(a)
myList <- list(a = 7, b = 5)
myList$a <- NULL  # works for data frames too
```

`NA` can hold a place but `NULL` cannot.
`NULL` is useful for having a function argument default to 'nothing'. See `help(crossprod)`


# Logical vectors
```{r}
answers <- c(TRUE, TRUE, FALSE, FALSE)
update <- c(TRUE, FALSE, TRUE, FALSE)

answers & update
answers | update
# note the vectorized boolean arithmetic

# what am I doing here?
identical(answers & update, as.logical(answers*update))
identical(answers | update, as.logical(answers + update))

sum(answers)
answers + update
```

# Data frames

- Data frames are combinations of vectors of the same length, but can be of different types
- Data frames are what is used for standard rectangular (record by field) datasets, similar to a spreadsheet
- Data frames are a functionality that both sets R aside from some languages (e.g., Matlab) and provides functionality similar to some statistical packages (e.g., Stata, SAS)

```{r}
require(foreign)
data=read.dta('../data/2004_labeled_processed_race.dta')
class(data)
head(data)
lapply(data, class)  # ignore this syntax for the moment and focus on the result
str(data)
```

# Data frames are (special) lists!

```{r}
is.list(data)
length(data)
data[[3]][1:5]
```

*lapply()* is a function used on lists, but it works here to apply the *class()* function to each element of the list, which in this case is each field/column.

# But lists are also vectors!

```{r}
length(data)
someFields <- data[c(3,5)]
head(someFields)
identical(data[c(3,5)], data[ , c(3,5)])
```


# Factors
- A factor is a very special and sometimes frustrating data type in R used for categorical data

```{r fac}
data=read.dta('../data/2004_labeled_processed_race.dta')
class(data$sizeplac)
head(data$sizeplac) # What order are the factors in?
levels(data[["sizeplac"]])
summary(data$sizeplac)
```

- What if we don't like the order these are in? Factor order is important for all kinds of things like plot type, regression output, and more

# Ordering the Factor
- Ordered factors simply have an additional attribute explaining the order of the levels of a factor
- This is a useful shortcut when we want to preserve some of the meaning provided by the order
- Think ordinal data

```{r orderedfac}
data <- within(data, 
     sizeplac_ord <- ordered(sizeplac, levels = levels(sizeplac)[c(5,3,4,2,1)])
)
head(data$sizeplac_ord)
levels(data$sizeplac_ord)
```

# Reclassifying Factors
- Turning factors into other data types can be tricky. All factor levels have an underlying numeric structure.

```{r fac2}
myfac <- factor(c('basic','proficient','advanced','basic', 'advanced', 'minimal'))
levels(myfac)
unclass(myfac)
```

- What's wrong?
- Be careful! The best way to convert a factor is to convert it to a character first.

```{r}
myfac <- factor(c('basic','proficient','advanced','basic', 'advanced', 'minimal'))
score = c(minimal = 3, basic = 1, advanced = 13, proficient = 7)
score["advanced"]
myfac[3]
score[myfac[3]]
score[as.character(myfac[3])]
```

What's going wrong?

# Subsetting

There are many ways to select subsets in R. The syntax below is useful for vectors, matrices, data frames, arrays and lists.

```{r}
vec <- rnorm(20)
mat <- matrix(vec, 4, 5)
rownames(mat) <- letters[1:4]
```
1) by direct indexing
```{r}
vec[c(3, 5, 12:14)]
vec[-c(3,5)]
mat[c(2,4), 5]
rowInd <- c(1, 3, 4)
colInd <- c(2, 2, 1)
mat[cbind(rowInd, colInd)]
```
2) by a vector of logicals
```{r}
cond <- vec > 0
vec[cond]
mat[mat[,1] > 0, ]
```
3) by a vector of names
```{r}
mat[c('a', 'd', 'a'), ]
```
4) using *subset()*
```{r}
subset(mtcars, mpg > 20)
```


# Assignment into subsets

We can assign into subsets by using similar syntax.

```{r}
vec[c(3, 5, 12:14)] <- 1:5
mat[2, ] <- rnorm(5)
mat[mat[,1] > 0, ] <- -Inf
```

# Strings

R has lots of functionality for character strings. Usually these are stored as vectors of strings, each with arbitrary length.

```{r}
chars <- c('hi', 'hallo', "mother's", 'father\'s', "He said, \"hi\"" )
length(chars)
nchar(chars)
paste("bill", "clinton", sep = " ")  # paste together a set of strings
paste(chars, collapse = ' ')  # paste together things in a vector
strsplit("This is the R bootcamp", split = " ")
grep("hi", chars)
gsub("hi", "Hi", chars)
substring(chars, 2, 3)
substring(chars, 2, 3) <- "ZZ"
```

# Regular expressions (regex or regexp)

Some of you may be familiar with using *regular expressions*, which is functionality for doing sophisticated pattern matching and replacement with strings. Python and Perl are both used extensively for such text manipulation. 

R has a full set of regular expression capabilities available through the *grep()*, *gregexpr()*, and *gsub()* functions (among others - many R functions will work with regular expressions).

You can basically do anything you need to in R, though the syntax may be a bit clunky at times.

# The working directory

To read and write from R, you need to have a firm grasp of where you are reading and writing from. 

```{r}
getwd()  # what directory will R look in?
setwd('~/Desktop/r-bootcamp-2013') # change the working directory
setwd('/Users/paciorek/Desktop/tmp') # absolute path
setwd('../r-bootcamp-2013') # relative path
```
Many errors and much confusion result from you and R not being on the same page in terms of where in the directory structure you are.

# Reading data into R

The workhorse for reading into a data frame is *read.table()*, which allows any separator (CSV, tab-delimited, etc.). *read.csv()* is a special case of *read.table()* for CSV files.

```{r}
data <- read.table("../data/RTAData.csv", sep = ",", head = TRUE)
head(data)
dim(data)
# great, we're all set, right?
# Not so fast...
unlist(lapply(data, class))
# ?read.table
data2 <- read.table("../data/RTAData.csv", sep = ",", head = TRUE, stringsAsFactors = FALSE)
data2[3,3]
unlist(lapply(data2, class))
# let's delve more deeply
levels(data[ , 2])[1:5]
data3 <- read.table("../data/RTAData.csv", sep = ",", head = TRUE, stringsAsFactors = FALSE, na.strings = c('NA', 'x'))
unlist(lapply(data3, class))
missing <- which(data[ , 2] == "")
data3[head(missing), ]
```

It's good to first look at your data in plain text format and then to check it after you've read it into R.

# Other ways to read data into R

The *read.table()* family of functions just skims the surface of things...

1) You can also read in a file as vector of characters, one character string per line of the file, and then post-process it. 
2) You can read fixed width format (constant number of characters per field) with *read.fwf()*

# Reading 'foreign' format data

We've already seen an example of reading data produced by another statistical package with `read.dta`. There are a number of other formats that we can handle for either reading or writing. Let's see `library(help = foreign)`.```

R can also read in (and write out) Excel files, netCDF files, HDF5 files, etc., in many cases through add-on packages from CRAN. 

A pause for a (gentle) diatribe:

Please try to avoid using Excel files as a data storage format. It's proprietary, complicated (can have multiple sheets), allows a limited number of rows/columns, and not easily readable/viewable (unlike simple text files).

# Writing data out from R

Here you have a number of options. 

1) You can write out R objects to an R Data file, as we've seen, using *save()* and *save.image()*.
2) You can use *write.csv()* and *write.table()* to write data frames/matrices to flat text files with delimiters such as comma and tab.
3) You can use *write()* to write out matrices in a simple flat text format.
4) You can use *cat()* to write to a file, while controlling the formatting to a fine degree.
5) The various file formats mentioned on the previous slide

# Writing out plots and tables
```{r}
pdf('myplot.pdf', width = 7, height = 7)
x <- rnorm(10); y <- rnorm(10)
plot(x, y)
dev.off()
```
`xtable()` formats tables for HTML and Latex (the default).

```{r}
require(xtable)
print(xtable(table(data$race, data$pres04)), type = "html")
```


# Breakout

read a tricky dataset in and manipulate it?
