% R bootcamp, Module 10: Advanced topics
% August 2013, UC Berkeley
% Chris Paciorek

```{r chunksetup, include=FALSE} 
# include any code here you don't want to show up in the document,
# e.g. package and dataset loading
library(foreign)
earnings <- read.dta('../data/heights.dta')
```

# Object-oriented programming in R

Note to BB: start recording.

Confusingly, R has three different systems for OOP, and none of them are as elegant and powerful as in Python or other languages more focused on OOP. That said, they get the job done for a lot of tasks.

* S3: informal system used for `lm()`, `glm()`, and many other core features in R in the *stats* package
* S4: more formal system, used with *lme4* 
* Reference Classes: new system allowing for passing objects by reference

# Basics of object-oriented programming

The basic idea is that coding is structured around *objects*, which have  *class* and *methods* that operate on objects in the class.

So basically like a list but with methods that are specifically associated with particular classes, as we've seen with the `lm` class.

Whether one needs to carefully specify the definition of the class (what the fields/components and methods are) is a distinguishing feature between S3 and S4.

# Working with S3 classes and methods

```{r}
library(methods)
yb <- sample(c(0, 1), 10, replace = TRUE)
yc <- rnorm(10)
x <- rnorm(10)
mod1 <- lm(yc ~ x)
mod2 <- glm(yb ~ x, family = binomial)
mod2$residuals # access field with list-like syntax

class(mod2)
is(mod2, "lm")
is.list(mod2)
names(mod2)

methods(class = "glm")

methods(predict)

print

# print.glm
```

When `print()` is called on a GLM object, it first calls the generic `print()`, which then recognizes that the first argument is of the class *glm* and immediately calls the right class-specific method.

# Making your own S3 class/object/method

Making an object and class-specific methods under S3 is simple. 

```{r}
rboot2013 <- list(month = 'August', year = 2013, instructor = 'Paciorek', attendance = 100)
class(rboot2013) <- "workshop"

rboot2013
is(rboot2013, "workshop")
rboot2013$instructor 

print.workshop <- function(x) {
    with(x,
       cat("A workshop held in ", month, " ", year, "; taught by ", instructor, ".\nThe attendance was ", attendance, ".\n", sep = ""))
    invisible(x)
}
rboot2013
```
 
Note that we rely on the generic `print()` already existing in R. Otherwise we'd need to create it.

# Using S4 classes and methods

Unlike S4, S4 classes have a formal definition and objects in the class must have the specified fields/components. 

The components of an S4 class are called slots. Instead of `x$field` you do `x@field`.

Here's a bit of an example of an S4 class for a linear mixed effects model from *lme4*:

```{r}
fm1 <- lmer(Reaction ~ Days + (Days|Subject), sleepstudy)
class(fm1)
methods("mer")
slotNames(fm1)
fm1$ranef
fm1@ranef
```

# A brief mention of Reference Classes

Reference classes are new in the last few years; for more information, see `?ReferenceClasses`.

Reference classes are somewhat like S4 in that a class is formally defined and R is careful about the fields of the class.

Methods for reference classes can modify the contents of fields in an object invisibly without the need to copy over the entire object.


# Computing on the language

One of the powerful capabilities you have in R is the ability to use R to modify and create R code. 

First we need to understand a bit about how R code is stored and manipulated when we don't want to immediately evaluate it.

When you send some code to R to execute, it has to 'parse' the input; i.e., to process it so that it know how to evaluate it. The parsed input can then be evaluated in the proper context (i.e., the right frame). 

We can capture parsed code before it is evaluated, manipulate it, and execute the modified result.

# Capturing and evaluating parsed code

```{r}
code <- quote(n <- 100)
code
class(code)
eval(code)
results <- rep(0, n)
moreCode <- quote(for(i in 1:n) {
    tmp <- rnorm(30)
    results[i] <- min(tmp)
})
class(moreCode)

newN <- 200
codeText <- paste("n", "<-", newN)
codeFromText <- parse(text = codeText)
eval(codeFromText)
n
```

So you could use R's string manipulation capabilities to write and then evaluate R code. Meta.

# Using R to automate working with object names 

Suppose you were given a bunch of objects named "x1", "x2", "x3", ... and you wanted to write code to automatically do some computation on them.

```{r}
x1 <- rnorm(5)
x2 <- rgamma(10)
x3 <- runif(20)
nVals <- 3
results <- rep(0, nVals)
for(i in 1:nVals) {  
  tmp <- eval(as.name(paste("x", i, sep = "")))
  # tmp <- get(paste("x", i, sep = "")) # an alternative
   results[i] <- mean(tmp)
}
```

Or suppose you needed to create "x1", "x2", "x3", automatically.

```{r}
nVals <- 3
results <- rep(0, nVals)
for(i in 1:nVals) {  
   assign(paste("x", i, sep = ""), rnorm(10))
}
x2
```

# Error and warning messages

When you write your own functions, and particularly for distributing to others, it's a good idea to:

* Check for possible errors (particularly in the input arguments) and give the user an informative error message
* Warn them if you're doing something they might not have anticipated

We can use `stop()` and `warning()` to do this. They're the same functions that are being called when you see an error message or a warning in reaction to your own work in R.

```{r}
mysqrt <- function(x) {
  if(is.list(x)) {
    warning("x is a list; converting to a vector")
    x <- unlist(x)
  }
  if(!is.numeric(x)) {
    stop("What is the square root of 'bob'?")
  } else {
      if(any(x < 0)) {
        warning("mysqrt: found negative values; proceeding anyway")
        x[x < 0] <- NaN
        x[x >= 0] <- (x[x >= 0])^(1/2)
        return(x)
      } else return(x^(1/2))
  }
}

mysqrt(c(1, 2, 3))
mysqrt(c(5, -7))
mysqrt(c('asdf', 'sdf'))
mysqrt(list(5, 3, 'ab'))
sqrt(c(5, -7))
sqrt('asdf')
sqrt(list(5, 3, 2))
```

So we've done something similar to what `sqrt()` actually does in R.

# 'Catching' errors

When you automate analyses, sometimes an R call will fail. But you don't want all of your analyses to grind to a halt because one failed. Rather, you want to catch the error, record that it failed, and move on.

For me this is most critical when I'm doing stratified analyses or sequential operations.

The `try()` function is a powerful tool here.

# Why we need to `try()`

Suppose we tried to do a stratified analysis of earnings on height within education levels. I'm going to do this as a for loop for pedagogical reasons, but again, it would be better to do this with apply/by/plyr type tools.


```{r}
mod <- list()
for(edLevel in unique(earnings$ed)) {
print(edLevel)
            sub <- subset(earnings, ed == edLevel)
            mod[[edLevel]] <- lm(earn ~ height, data = sub)
}
```

1) What happened?
2) Why did it go through the education levels in that order? (12, 16, 17, ...)

# How we can `try()` harder

```{r}
mod <- list()
for(edLevel in unique(earnings$ed)) {
print(edLevel)
            sub <- subset(earnings, ed == edLevel)
            tmp <- try(lm(earn ~ height, data = sub))
            if(is(tmp, "try-error")) mod[[edLevel]] <- NA else mod[[edLevel]] <- tmp
}
mod[[2]]
mod[[3]]
```


# File encodings

Text (either in the form of a file with regular language in it or a data file with fields of character strings) will often contain characters that are not part of the [limited ASCII set of characters](http://en.wikipedia.org/wiki/ASCII), which has 128 characters and control codes; basically what you see on a standard US keyboard.

UTF-8 is an encoding for the Unicode characters that include more than 110,000 characters from 100 different alphabets/scripts. It's widely used on the web.

Latin-1 encodes a small subset of Unicode and contains the characters used in many Europeon languages (e.g., letters with accents).

# Dealing with encodings in R

To read files with other characters correctly into R, you may need to tell R what encoding the file is in. E.g., see help on `read.table()` for the *fileEncoding* and *encoding* arguments. 

With strings already in R, you can convert between encodings:
```{r}
text <- "_Melhore sua seguran\xe7a_"
iconv(text, from = "latin1", to = "UTF-8")
iconv(text, from = "latin1", to = "ASCII", sub = "???")
```

You can mark a string with an encoding so R can display it correctly:
```{r}
x <- "fa\xE7ile"
Encoding(x) <- "latin1"
x
```

# Line endings in text files

Windows, Mac, and Linux handle line endings in text files somewhat differently. So if you read a text file into R that was created in a different operating system you can run into difficulties.

* In Windows lines end in both a newline (the ASCII character \n) and a carriage return (\r). 
* In UNIX, lines end in only a newline.
* On a Mac, lines may end only in a carriage return.

So in UNIX you might see `^M` at the end of lines when you open a Windows file in a text editor. The dos2unix or fromdos commands can do the necessary conversion

In Windows you might have a UNIX text file appear to be all one line. The *unix2dos* or *todos* commands in UNIX can do the conversion. 

There is a UNIX utility call *mac2unix* that can convert Mac text files.

There may also be Windows tools to deal with this. 

# Working with databases

R has the capability to read and write from a variety of relational database management systems (DBMS). Basically a database is a collection of rectangular format datasets (tables). Some of these tables have fields in common so it makes sense to merge (i.e., join) information from multiple tables. E.g., you might have a database with a table of student information, a table of teacher information and a table of school information. 

The *DBI* package provides a front-end for manipulating databases from a variety of DBMS (MySQL, SQLite, Oracle, among others)

Basically, you tell the package what DBMS is being used on the back-end, link to the actual database, and then you can use the standard functions in the package regardless of the back-end. 

# Database example

The Current Index to Statistics contains records of article and author information from most Statistics journals. 

The database is not freely available, so it's not on the course bSpace or Github, but I can do a demo here.

```{r}
library(RSQLite)  # DBI is a dependency
db <- dbConnect(SQLite(), dbname = "../data/cis.db") # cis.db is an SQLite database
dbListTables(db)
dbListFields(db, "articles")
breiman <- dbGetQuery(db, "select * from authors where name like 'Breiman%'")
```

# Breakout

Have a snack!
